Software architectures to integrate workflow engines in science gateways

### 2016-05-15 paper.tex
Minor updates

In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools among which
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and
WS-VLAM~\cite{wsvlam}. \closednote{vladimir}{As a past developer of it, I
  would suggest to refer to WS-VLAM~\cite{wsvlam} as well, despite it
  is not very famous, but it played an important role for Amsterdam
  e-science workflows}{Tristan}{Done, thanks.} Such workflow engines usually describe
applications in a high-level language with specific data and control
flow constructs, parallelization operators, visual edition tools,
links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

@@

The pool architecture is overall the most scalable, which is not
surprising since it is designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however,
should not conceal the unique characteristics of architectures regarding
this criterion. Nested workflows are the only architecture that can
easily accommodate distributed workflow execution, which can be
critical in some cases. At the same time, the scheduling constraints
created by task encapsulation and nested workflows may become problematic  depending on the type of targeted
infrastructure. Non-reliable infrastructures,
for example, could hardly cope with workflow engines being wrapped in
computing tasks as done in task encapsulation. The availability of
multiple engine instances could also become a critical feature for
science gateways with important workloads, which would favor task
encapsulation and pool. \note{Naj}{I feel this paragraph is a bit
  vague, and doesn't distill and contextualize the thorough work
  described before.}

Differences in other specific features should not be neglected. Task encapsulation, nested
workflows (with service invocation) and workflow conversion (with service
invocation) are the only architectures that intrinsically support meta-workflows. \closednote{vladimir}{As mentioned earlier by Silvia, this might not be exactly the case, depending on the point of view}{Tristan}{The example given by Silvia (WS-PGRADE) actually comes down to the nested workflow architecture. I added 'intrinsically' to the sentence to better reflect that.} Besides,
task encapsulation and nested workflows may hamper
fine-grained debugging. 
The availability of fine-grained debugging information may also be
critical to efficient user support.

@@

The abstract nested workflows and workflow conversion patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with task encapsulation,
could also be envisaged. We chose to limit ourselves to instantiations
with service invocation because the resulting architectures are
implemented in real systems, and because service invocation is largely
used. Nevertheless, it could be interesting to explore other types of
instantiations, in particular nested workflows with task
encapsulation.\answerednote{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}{Tristan}{Good idea, will think about it.}

The particular technical or historical context of a science gateway
project may obviously influence the choice of an architecture to
integrate workflow engines. For instance, some workflow engines may be
already available as web services\note{Naj}{This is a good place to offer some examples}, which would tend to favor service
invocation, and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

@@

The early work~\cite{yuTaxonomy} proposes a generic architecture for
grid workflow systems that is based on the workflow reference model
defined by the Workflow Management
Coalition\footnote{\url{http://www.wfmc.org}}.  Deelman et
al. \cite{deelman2009workflows} have further characterized the
features of workflow systems and proposed a general architecture??
These two works, however, do not mention science gateways.

Shahand et al.~\cite{shahand:2015ab} presents the Science Gateway
Canvas, which is a business reference model where the most relevant
functional components are organized into functional groups. Although
workflow management is mentioned as a possible component to coordinate
distributed computations in science gateways, the paper does not
comment on architecture to achieve this.

Olabarriaga et al.~\cite{olabarriaga2014} presents a user-centered
view of the ecosystem of science gateways with focus on workflow
management. Their proposed layered system architecture includes
gateway, workflow engine and distributed infrastructure components,
similarly to the "service invocation" pattern. However other patterns
are not identified.
\note{vladimir}{@silvia, does it make sense to mention workflow classification efforts based on workflow motifs and so on?}

>>>>

\label{sec:introduction-engines}
In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools among which
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and
WS-VLAM~\cite{wsvlam}. \closednote{vladimir}{As a past developer of it, I
  would suggest to refer to WS-VLAM~\cite{wsvlam} as well, despite it
  is not very famous, but it played an important role for Amsterdam
  e-science workflows}{Tristan}{Done, thanks.} Such workflow engines usually describe
applications in a high-level language with specific data and control
flow constructs, parallelization operators, visual edition tools,
links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

@@

The pool architecture is overall the most scalable, which is not
surprising since it is designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however,
should not conceal the unique characteristics of architectures regarding
this criterion. Nested workflows are the only architecture that can
easily accommodate distributed workflow execution, which can be
critical in some cases. At the same time, the scheduling constraints
created by task encapsulation and nested workflows may become problematic  depending on the type of targeted
infrastructure. Non-reliable infrastructures,
for example, could hardly cope with workflow engines being wrapped in
computing tasks as done in task encapsulation. The availability of
multiple engine instances could also become a critical feature for
science gateways with important workloads, which would favor task
encapsulation and pool. \closednote{Naj}{I feel this paragraph is a bit
  vague, and doesn't distill and contextualize the thorough work
  described before.}{Tristan}{Rejected comment :)}

Differences in other specific features should not be neglected. Task encapsulation, nested
workflows (with service invocation) and workflow conversion (with service
invocation) are the only architectures that intrinsically support meta-workflows. \closednote{vladimir}{As mentioned earlier by Silvia, this might not be exactly the case, depending on the point of view}{Tristan}{The example given by Silvia (WS-PGRADE) actually comes down to the nested workflow architecture. I added 'intrinsically' to the sentence to better reflect that.} Besides,
tight integration is the only architecture that allows
fine-grained debugging. 
The availability of fine-grained debugging information may be
critical to efficient user support.

@@

The abstract nested workflows and workflow conversion patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with task encapsulation,
could also be envisaged. We chose to limit ourselves to instantiations
with service invocation because the resulting architectures are
implemented in real systems, and because service invocation is largely
used. Nevertheless, it could be interesting to explore other types of
instantiations.\closednote{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}{Tristan}{Not sure how to do this without making the conclusion sound too specific to these systems.}

The particular technical or historical context of a science gateway
project may obviously influence the choice of an architecture to
integrate workflow engines. For instance, many workflow engines are
already available as web services\closednote{Naj}{This is a good place to offer some examples}{Tristan}{Many of them are, precised the text, not sure adding explicit examples makes sense at this stage.}, which tends to favor service
invocation, and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

@@

There is abundant literature describing specific workflow engines and
systems; some of the main references are cited in
Section~\ref{sec:introduction-engines}. A few works described the
architecture and features of workflow system at an abstract level. For
instance, the early work~\cite{yuTaxonomy} proposes a generic
architecture for grid workflow systems that is based on the workflow
reference model defined by the Workflow Management
Coalition\footnote{\url{http://www.wfmc.org}}.  Deelman et
al. \cite{deelman2009workflows} have further characterized the
features of workflow systems. These two works, however, do not mention
science gateways.

Numerous science gateways have been described, as mentioned in
Section~\ref{sec:intro_sg}. However, only a few works focused on
science gateway architectures. Shahand et al.~\cite{shahand:2015ab}
presents the Science Gateway Canvas, which is a business reference
model where the most relevant functional components are organized into
functional groups. Although workflow management is mentioned as a
possible component to coordinate distributed computations in science
gateways, the paper does not comment on architecture to achieve this.

Olabarriaga et al.~\cite{olabarriaga2014} presents a user-centered
view of the ecosystem of science gateways with focus on workflow
management. Their proposed layered system architecture includes
gateway, workflow engine and distributed infrastructure components,
similarly to the ``service invocation'' pattern. However other patterns
are not identified.

\closednote{vladimir}{@silvia, does it make sense to mention workflow classification efforts based on workflow motifs and so on?}{Tristan}{I'm  not  sure. Which papers are you thinking of?}
